Symmetric Circuits for Rank Gates

### 2018-03-28 sections/translating-formulas-to-cicuits.tex
A complete draft minus the conclusion. Large changes made to a number of sections.
I have made changes to the section "symmetric circuits" and updated every other section in line with changes made.
minor revisions to Sections 6 and 7

\begin{lem}
  There is a function that takes as input a $\FOrk$-formula $\theta(\vec{x})$
  and $n \in \nats$ and outputs a circuit $C$ that translates $\theta(\vec{x})$
  for $n$. Moreover, this function is computable and there is a polynomial in
  $p$ such that for an input $(\theta(\vec{x}), n)$ the algorithm computing this
  function terminates in at most $p(n^{\width(\theta)}\vert \cl{\theta} \vert)$
  many steps.
  \label{lem:translating-FOrk}
\end{lem}
\begin{proof}
  Let $\tau$ be the vocabulary of $\theta(\vec{x})$. Let $k := \width(\theta)$.
  Suppose $\tau$ is empty. Since $\tau$, the vocabulary of $\theta$, is empty,
  all of the atomic formulas in $\theta(\vec{x})$ are of the form $y_1 = y_2$,
  where $y_1$ and $y_2$ are variables. It can be shown that for $\vec{a} \in
  n^k$, we can evaluate $[n] \models \theta[\alpha^{\vec{x}}_{\vec{a}}]$ in time
  polynomial in $n^{k} \vert \cl{\theta} \vert$. Let $f : [n]^k \rightarrow
  \{0,1\}$ be defined such that $f (\vec{a}) = 1$ if, and only if, $[n] \models
  \theta[\alpha^{\vec{x}}_{\vec{a}}]$. Let $C := \langle G, \Sigma, \Omega,
  \Lambda, L \rangle$, where
  \begin{itemizens}
  \item $G = \{g_0, g_1\} \cup \{g_{\vec{a}} : \vec{a} \in [n]^{k}\}$,
  \item $\Sigma (g_0) = 0$, $\Sigma (g_1) = 1$ and $\Sigma (g_{\vec{a}}) =
    \land[1]$ for all $\vec{a} \in [n]^k$,
  \item $\Omega : [n]^k \rightarrow G$ is defined by $\Omega(\vec{a}) =
    g_{\vec{a}}$ for all $\vec{a} \in [n]^k$,
  \item $\Lambda = \emptyset$, and
  \item for $\vec{a} \in [n]^k$, $L(g_{\vec{a}}) : [1] \rightarrow G$ is defined
    by $L(g_{\vec{a}})(1) = g_{f(\vec{a})}$.
  \end{itemizens}

@@

  Suppose $\tau$ is not empty. Let $R \in \tau$. We first define two helper
  formulas. For a variable $y$ let $\op{no-op}(y) := (R(y, y) \lor (\neg R(y,
  y)))$. For a sequence of variables $\vec{y} = (y_1, \ldots, y_m)$ let
  $\op{tag} (\vec{y}) := (\op{no-op}(y_1) \land (\op{no-op}(y_2) \land (
  \op{no-op}(y_2) \land ( \ldots \land (\op{no-op}(y_m)) \ldots))))$. Let
  $\lambda (\vec{x})$ be a formula constructed from $\theta(\vec{x})$ by
  replacing each sub-formula $\psi(\vec{y})$ of the form $\rank^r_p \vec{w}
  \vec{z} . \phi$ with the formula $\rank^r_p \vec{w}\vec{z} . ((\forall u . u =
  u) \land \phi) \land \op{tag}(\vec{w} \cup \vec{z})$. Since this process
  involves inserting conjunctions of tautologies, the matrix defined at point
  remains unchanged (i.e. $\phi \Leftrightarrow ((\forall u . u = u) \land \phi)
  \land \op{tag}(\vec{w} \cup \vec{z})$ is a tautology), and so the evaluation
  of $\psi(\vec{y})$ is unchanged by this replacement. It follows that $\lambda
  (\vec{x})$ and $\theta (\vec{x})$ are equivalent. The intuition is that
  $\op{tag}(\vec{w} \cup \vec{z})$ appends a tower of conjunctions of
  tautologies, with each tautology referencing a unique variable from $\vec{w}
  \cup \vec{z}$. When we construct the circuit, this tower of tautologies will
  act to `tag' each input to the rank gate with a unique gadget. We then show
  that the presence of these gadgets ensures that no two inputs to a rank gate
  are syntactically-equivalent, and so the circuit is
  transparent.

@@

  \begin{itemizens}
  \item If $\psi[\alpha] = T(\vec{a})$ for some $T \in \tau$ and $\vec{a} \in
    \arty(T)$, then $\Sigma (g) = T$ and $\Lambda_T (g) = \vec{a}$.
  \item If $\psi[\alpha]$ is $0$ or $1$, then $\Sigma (g) = \psi[\alpha]$.
  \item If $\psi$ is of the form $\rank^r_p \vec{w} \vec{z}. \phi$, let $A =
    [n]^{\vert \vec{w}\vert}$ and $B = [n]^{\vert \vec{z}\vert}$, and $\Sigma
    (g_{\psi, \alpha}) = \rank[A, B]$. Let $L(g) : A \times B \rightarrow G$ be
    defined by $L(g)(\vec{a},\vec{b}) = g_{\phi[\beta]}$, where $\beta \in
    [n]^{\free{\phi}}$, and where $\alpha$, and the assignments
    $\alpha^{\vec{w}}_{\vec{a}}$ and $\alpha^{\vec{z}}_{\vec{b}}$, are
    compatible with $\beta$.
  \item If $\psi$ is of the form $Q z . \phi(\vec{y}, z)$, for $Q \in \{\forall,
    \exists\}$, then if $Q = \forall$ we have $\Sigma (g) = \land [n]$,
    otherwise $\Sigma (g) = \lor [n]$. Let $L(g) : [n] \rightarrow G$ be defined
    for $i \in [n]$ by $L(g)(i) = g_{\phi[\beta_i]}$, where $\beta_i \in
    [n]^{\free{\psi}}$, and where both $\alpha$ and $\alpha^{z}_{i}$ are
    compatible with $\beta_i$.
  \item If $\psi$ is of the form $\phi_1 \land \phi_2$, then $\Sigma(g) =
    \land[2]$ and $L(g) : [2] \rightarrow G$ is defined for $i \in [2]$ by
    $L(g)(i) = g_{\phi_i[\beta_i]}$, where $\beta_i \in [n]^{\free{\psi}}$ and
    $\beta_i$ is compatible with $\alpha$. The same approach is used in the
    disjunctive case.
  \item if $\psi$ is of the form $\neg \phi$ then $\Sigma (g) = \neg$ and $L(g):
    [1] \rightarrow G$ is defined by $L(g)(1) = g_{\phi[\alpha]}$.
  \end{itemizens}

@@

We notice that $\vert C \vert \leq n^{\width(\lambda)} \vert \cl{\lambda} \vert
\leq n^{\width(\theta)} \vert \cl{\theta} \vert (5 + \width(\theta))$. It is not
hard to see that $C$ (and $\lambda(\vec{x})$) can be constructed in a number of
steps bounded by a polynomial in the size of $C$. This completes the proof of
the result.

@@

@@

\bibliography{../references.bib}
% \ifcsdef{mainfile}{}{\bibliography{../references.bib}}
\end{document}

>>>>

\begin{lem}
  There is a function that takes as input a $\FOrk$-formula $\theta(\vec{x})$
  and $n \in \nats$ and outputs a circuit $C$ that translates $\theta(\vec{x})$
  for $n$. Moreover, this function is computable and there is a polynomial in
  $p$ such that for an input $(\theta(\vec{x}), n)$ the algorithm computing this
  function terminates in at most $p(n^{\width(\theta)}\vert \cl{\theta} \vert)$
  many steps.
  \label{lem:translating-FOrk}
\end{lem}
\begin{proof}
  Let $\tau$ be the vocabulary of $\theta(\vec{x})$. Let $k := \width(\theta)$.
  Suppose first that $\tau$ is empty, so
  all of the atomic formulas in $\theta(\vec{x})$ are of the form $y_1 = y_2$,
  where $y_1$ and $y_2$ are variables.  It can be shown that for $\vec{a} \in
  n^k$, we can evaluate $[n] \models \theta[\alpha^{\vec{x}}_{\vec{a}}]$ in time
  polynomial in $n^{k} \vert \cl{\theta} \vert$. Let $f : [n]^k \rightarrow
  \{0,1\}$ be defined such that $f (\vec{a}) = 1$ if, and only if, $[n] \models
  \theta[\alpha^{\vec{x}}_{\vec{a}}]$. Let $C := \langle G, \Sigma, \Omega,
  \Lambda, L \rangle$, where
  \begin{myitemize}
  \item $G = \{g_0, g_1\} \cup \{g_{\vec{a}} : \vec{a} \in [n]^{k}\}$,
  \item $\Sigma (g_0) = 0$, $\Sigma (g_1) = 1$ and $\Sigma (g_{\vec{a}}) =
    \AND[1]$ for all $\vec{a} \in [n]^k$,
  \item $\Omega : [n]^k \rightarrow G$ is defined by $\Omega(\vec{a}) =
    g_{\vec{a}}$ for all $\vec{a} \in [n]^k$,
  \item $\Lambda = \emptyset$, and
  \item for $\vec{a} \in [n]^k$, $L(g_{\vec{a}}) : [1] \rightarrow G$ is defined
    by $L(g_{\vec{a}})(1) = g_{f(\vec{a})}$.
  \end{myitemize}

@@

  Suppose now $\tau$ is not empty.  We first define a transformation
  of the formula $\theta$ to an equivalent formula $\lambda$ which has
  the property that the \emph{natural} translation of $\lambda$ to a
  circuit (given later) is transparent.
Let $R \in \tau$.  We first define two helper
  formulas. For a variable $y$ let $\op{no-op}(y) := (R(y, y) \lor (\neg R(y,
  y)))$. For a sequence of variables $\vec{y} = (y_1, \ldots, y_m)$ let
  $\op{tag} (\vec{y}) := (\op{no-op}(y_1) \land (\op{no-op}(y_2) \land (
  \op{no-op}(y_2) \land ( \ldots \land (\op{no-op}(y_m)) \ldots))))$.
  Now, let  $\lambda (\vec{x})$ be the formula constructed from $\theta(\vec{x})$ by
  replacing each sub-formula $\psi(\vec{y})$ of the form $\rank^r_p \vec{w}
  \vec{z} . \phi$ with the formula $\rank^r_p \vec{w}\vec{z} . ((\forall u . u =
  u) \land \phi) \land \op{tag}(\vec{w} \cup \vec{z})$.  Since we
  always replace a subformula $\phi$ with a logically equivalent
  formula, it follows that $\lambda
  (\vec{x})$ and $\theta (\vec{x})$ are equivalent. The intuition is that
  $\op{tag}(\vec{w} \cup \vec{z})$ appends a tower of conjunctions of
  tautologies, with each tautology referencing a unique variable from $\vec{w}
  \cup \vec{z}$. When we construct the circuit, this tower of tautologies will
  act to `tag' each input to the rank gate with a unique gadget. We then show
  that the presence of these gadgets ensures that no two inputs to a rank gate
  are syntactically-equivalent, and so the circuit is
  transparent.

@@

  \begin{myitemize}
  \item If $\psi[\alpha] = T(\vec{a})$ for some $T \in \tau$ and $\vec{a} \in
    \arty(T)$, then $\Sigma (g) = T$ and $\Lambda_T (g) = \vec{a}$.
  \item If $\psi[\alpha]$ is $0$ or $1$, then $\Sigma (g) = \psi[\alpha]$.
  \item If $\psi$ is of the form $\rank^r_p \vec{w} \vec{z}. \phi$, let $A =
    [n]^{\vert \vec{w}\vert}$ and $B = [n]^{\vert \vec{z}\vert}$, and $\Sigma
    (g_{\psi, \alpha}) = \rank[A, B]$. Let $L(g) : A \times B \rightarrow G$ be
    defined by $L(g)(\vec{a},\vec{b}) = g_{\phi[\beta]}$, where $\beta \in
    [n]^{\free{\phi}}$, and where $\alpha$, and the assignments
    $\alpha^{\vec{w}}_{\vec{a}}$ and $\alpha^{\vec{z}}_{\vec{b}}$, are
    compatible with $\beta$.
  \item If $\psi$ is of the form $Q z . \phi(\vec{y}, z)$, for $Q \in \{\forall,
    \exists\}$, then if $Q = \forall$ we have $\Sigma (g) = \AND[n]$,
    otherwise $\Sigma (g) = \OR[n]$. Let $L(g) : [n] \rightarrow G$ be defined
    for $i \in [n]$ by $L(g)(i) = g_{\phi[\beta_i]}$, where $\beta_i \in
    [n]^{\free{\psi}}$, and where both $\alpha$ and $\alpha^{z}_{i}$ are
    compatible with $\beta_i$.
  \item If $\psi$ is of the form $\phi_1 \land \phi_2$, then $\Sigma(g) =
    \AND[2]$ and $L(g) : [2] \rightarrow G$ is defined for $i \in [2]$ by
    $L(g)(i) = g_{\phi_i[\beta_i]}$, where $\beta_i \in [n]^{\free{\psi}}$ and
    $\beta_i$ is compatible with $\alpha$. The same approach is used in the
    disjunctive case.
  \item if $\psi$ is of the form $\neg \phi$ then $\Sigma (g) = \neg$ and $L(g):
    [1] \rightarrow G$ is defined by $L(g)(1) = g_{\phi[\alpha]}$.
  \end{myitemize}

@@

We note that $\vert C \vert \leq n^{\width(\lambda)} \vert \cl{\lambda} \vert
\leq n^{\width(\theta)} \vert \cl{\theta} \vert (5 + \width(\theta))$. It is not
hard to see that $C$ (and $\lambda(\vec{x})$) can be constructed in a number of
steps bounded by a polynomial in the size of $C$. This completes the proof of
the result.

@@

  \begin{remark}
    A: Need more specific references than the above.
  \end{remark}

@@

% \ifcsdef{mainfile}{}{\bibliography{../references.bib}}
\end{document}
